Distortion-corrected kernel density estimate for Riemannian manifolds

### 2022-07-06 paper/paper.tex
Change title

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Kernel density estimate for embedding of Riemannian manifolds},
            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\title{Kernel density estimate for embedding of Riemannian manifolds}

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. \textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with the lowest density estimates. The proposed schematic is shown in Figure \ref{fig:vkde}. The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.

@@

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).

\hypertarget{dckde}{%
\subsection{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}

However, in application, we usually find the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\). Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\). Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation}
where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).

@@

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Distortion-corrected kernel density estimate for Riemannian manifolds},
            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\title{Distortion-corrected kernel density estimate for Riemannian manifolds}

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. \textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with the lowest density estimates.

@@

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.

\hypertarget{dckde}{%
\subsection{Distortion-corrected kernel density estimator for manifold embedding}\label{dckde}}

However, in application, we usually find the observations lying on a \(d\)-dimensional manifold embedded in a much higher \(s\)-dimensional space, where \(d \ll s\). Instead of sampling from the undering manifold \(M\), we apply manifold learning algorithms to reduce dimension and find a \(d\)-dimensional representation of the manifold \(M\).
Notice that this representation is not necessarily isometry to \(M\), and the underlying manifold is usually unknown in real data sets. Then the density estimator in Equation \eqref{eq:denriem} proposed by \textcite{Pelletier2005-vu} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
Motivated on the estimator in \eqref{eq:denriem}, we propose a distortion-corrected kernel density estimator (DC-KDE) of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r} \frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|} K\bigg( \frac{(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})}{r} \bigg),
\end{equation}
where \(h(\pmb{y_i}), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and \(h(x)\) is positive definite with dimension \(d\times d\), \(\frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})\) is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point \(\pmb{y_i} \ \mathbb{R}^d\). Then the only problem left is the estimation of the Riemannian metric at each embedding point.

@@

The proposed schematic is shown in Figure \ref{fig:vkde}. The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
